---
title: "Practical Machine Learning Project"
author: "Katie T"
date: "8/19/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Executive Summary
In this analysis, we will be building a model to predict the manner (classe) in which a user exercises using data of accelerometers on the belt, forearm, arm, and dumbbell of 6 participants. We tested three different methods(random forest, gradient boosting machine, and linear discriminant analysis) and found our random forest model preforms best. Finally we applied this model to the data set pml-testing to predict the manner in which the users performed the exercises.

## Exploratory Analysis

Graph 1 shows the frequency of the classe variable. From this we see that A is the most common outcome for classe in our data set. 

```{r include=FALSE}

library(caret)
```

```{r echo=FALSE}
set.seed(125)
#Load Data
pml <- read.csv("C:\\Users\\kathe\\OneDrive\\Documents\\Coursera\\Practical_Machine_Learning\\Project\\pml-training.csv")


plot(as.factor(pml$classe),main='Graph 1: Classe Frequency',xlab='Classe',ylab='Frequency' )
```

Further we see that the aggregated variables are mostly missing so we will remove these for the analysis. This leaves the following columns for the analysis:
```{r echo=FALSE}


cleandata <- pml[,
                 c
                 ("roll_belt","pitch_belt","yaw_belt","total_accel_belt","gyros_belt_x","gyros_belt_y"
                   ,"gyros_belt_z","accel_belt_x","accel_belt_y","accel_belt_z","magnet_belt_x","magnet_belt_y"
                   ,"magnet_belt_z","roll_arm","pitch_arm","yaw_arm","total_accel_arm","gyros_arm_x"
                   ,"gyros_arm_y","gyros_arm_z","accel_arm_x","accel_arm_y","accel_arm_z","magnet_arm_x"
                   ,"magnet_arm_y","magnet_arm_z","roll_dumbbell","pitch_dumbbell","yaw_dumbbell"
                   ,"total_accel_dumbbell","gyros_dumbbell_x","gyros_dumbbell_y","gyros_dumbbell_z"
                   ,"accel_dumbbell_x","accel_dumbbell_y","accel_dumbbell_z","magnet_dumbbell_x"
                   ,"magnet_dumbbell_y","magnet_dumbbell_z","roll_forearm","pitch_forearm","yaw_forearm"
                   ,"total_accel_forearm","gyros_forearm_x","gyros_forearm_y","gyros_forearm_z"
                   ,"accel_forearm_x","accel_forearm_y","accel_forearm_z","magnet_forearm_x"
                   ,"magnet_forearm_y","magnet_forearm_z","classe")]
colnames(cleandata)

```
## Method 

We create a training and testing sets with the training set consisting of 70% of the data set. We will use cross-validation within model to uses to compare the models. The test set will be used to do a final validation on the chosen model. Next, we preprocess the data using principal component analysis with p=.7. We will then use this data to run the three models as well as use the same method to preprocess the testing set. We  also use trainControl function with method="cv"and number=10 in order to use in each of the models to preform cross validation.
```{r echo=FALSE}
trainSeg <- createDataPartition(cleandata$classe, p=.7,list=FALSE)
training<-cleandata[trainSeg,]
testing<-cleandata[-trainSeg,]
preProc<-preProcess(training, method="pca",.8)
trainPC<-predict(preProc, training)
testPC<-predict(preProc,testing)
ctrl<-trainControl(method="cv", number=10)
```


### Random Forest

We first create a random forest model using the new training set. 

```{r echo=FALSE}
modelFitrf<-train(classe~ ., method="rf", data=trainPC,trControl=ctrl)
modelFitrf$finalModel

```
### Gradient Boosting Machine

Next, we create a model using gradient boosting machine. 
```{r include=FALSE}
modelFitgbm<-train(classe~ .
                  , method="gbm", data=trainPC, trControl=ctrl)
```
```{r echo=FALSE}
modelFitgbm$finalModel

```
### Linear Discriminant Analysis
Finally we create a model using linear discriminant analysis.
```{r echo=FALSE}
modelFitlda<-train(classe~ .
                   , method="lda", data=trainPC, trControl=ctrl)
modelFitlda$finalModel

```
### Model Comparison and Evaluation

Looking at the confusion matrices from the cross-validation we see that random forest model has the highest accuracy and
thus the lowest out of sample error (1-accuracy). For this reason we will use the random forest model.

Random Forest:
```{r echo=FALSE}
confusionMatrix.train(modelFitrf)
```
Gradient Boosting Machine Test Set:
```{r echo=FALSE}
confusionMatrix.train(modelFitgbm)
```
Linear Discriminant Analysis Test Set:
```{r echo=FALSE}
confusionMatrix.train(modelFitlda)
```

### Model Evaluation

To evaluate our model we will run the model on the preprocessed test set. From the confusion matrix below we see that the model has a 96% accuracy on the test set with an out of sample error of 0.4. It also does a good job at predicting all the class types.
```{r echo=FALSE}
rf<-confusionMatrix(as.factor(testing$classe),as.factor(predict(modelFitrf$finalModel, testPC)))
rf
```

### Model Application 
Finally we will use the model to predict using the data set pml-testing which produces the following output:
```{r echo=FALSE}
pmltestcase <-read.csv("C:\\Users\\kathe\\OneDrive\\Documents\\Coursera\\Practical_Machine_Learning\\Project\\pml-testing.csv")

cleanpmltest <- pmltestcase[,
                 c
                 ("roll_belt","pitch_belt","yaw_belt","total_accel_belt","gyros_belt_x","gyros_belt_y"
                   ,"gyros_belt_z","accel_belt_x","accel_belt_y","accel_belt_z","magnet_belt_x","magnet_belt_y"
                   ,"magnet_belt_z","roll_arm","pitch_arm","yaw_arm","total_accel_arm","gyros_arm_x"
                   ,"gyros_arm_y","gyros_arm_z","accel_arm_x","accel_arm_y","accel_arm_z","magnet_arm_x"
                   ,"magnet_arm_y","magnet_arm_z","roll_dumbbell","pitch_dumbbell","yaw_dumbbell"
                   ,"total_accel_dumbbell","gyros_dumbbell_x","gyros_dumbbell_y","gyros_dumbbell_z"
                   ,"accel_dumbbell_x","accel_dumbbell_y","accel_dumbbell_z","magnet_dumbbell_x"
                   ,"magnet_dumbbell_y","magnet_dumbbell_z","roll_forearm","pitch_forearm","yaw_forearm"
                   ,"total_accel_forearm","gyros_forearm_x","gyros_forearm_y","gyros_forearm_z"
                   ,"accel_forearm_x","accel_forearm_y","accel_forearm_z","magnet_forearm_x"
                   ,"magnet_forearm_y","magnet_forearm_z")]
testcasePC<-predict(preProc,cleanpmltest)
testcase<-predict(modelFitrf$finalModel, testcasePC)
testcase
```
 
